---
title: "ST201 Data Analysis"
subtitle:  "Discrete Probability Distributions"
author: "Prof. Niamh Cahill (she/her)"
format: 
  revealjs:
    highlight-style: github
editor: visual
---

## Discrete Random Variables {style="font-size: 70%;"}

A **discrete random variable** represents outcomes that take on distinct integer values, often counted or listed.

### Example: Coin Tossing Experiment

Imagine tossing a coin **3 times**:

-   **Possible Outcomes**: \[ {HHH, HHT, HTH, THH, HTT, THT, TTH, TTT} \] There are **8 equally likely outcomes**.

-   Define ( X ): the number of heads observed.

### Properties of $X$

-   **Possible Values of** $X$:
    -   $X = 0, 1, 2,$ or $3$ (number of heads from 0 to 3).
-   **Interpretation**:
    -   $X$ is a discrete random variable because it counts a finite set of outcomes.

## Probability Mass Function {style="font-size: 70%;"}

A **Probability Mass Function (PMF)** is a function that provides the probability that a **discrete random variable** is exactly equal to a particular value.

For a discrete random variable $X$ with possible values $x_1, x_2, \dots, x_n$, the PMF, denoted as $P(X = x)$ or $f(x)$, assigns a probability to each possible outcome $x$ such that:

$$P(X = x) = f(x)$$

::: callout-note
## Properties of a Probability Mass Function:

1.  **Non-negativity**: $f(x) \geq 0$ for all $x$.
2.  **Normalization**: The sum of all probabilities equals 1, so $\sum_x f(x) = 1$.
:::

## Types of Discrete Distributions {style="font-size: 60%;"}

We'll explore **three important types** of discrete distributions, each with unique characteristics and applications:

1.  **Bernoulli Distribution**:
    -   **Definition**: Models a single trial with two possible outcomes, typically "success" (1) or "failure" (0).
    -   **Example**: Testing for a defect in a manufactured product. Each product can either be defective (1) or non-defective (0).
    -   **Parameter**: Probability of success, ( p ).
2.  **Binomial Distribution**:
    -   **Definition**: Represents the number of successes in a fixed number of independent Bernoulli trials.
    -   **Example**: Counting the number of defective products in a batch of 20 items inspected on an assembly line.
    -   **Parameters**: Number of trials, $n$, and probability of success, $p$.
3.  **Poisson Distribution**:
    -   **Definition**: Models the number of times an event occurs within a fixed interval of time or space, given the average rate of occurrence.
    -   **Example**: Number of emails received in an hour.
    -   **Parameter**: Average rate of occurrence, $\lambda$.

These distributions help model various types of **discrete events**.

## The Bernoulli Distribution {style="font-size: 70%;"}

A 'Bernoulli trial' or 'Bernoulli random variable' is where there are just two possible outcomes which we denote either a 'success' or a 'failure'.

**Example** X = Defect in a manufactured product. The product can either be defective (1) or non-defective (0).

**Probability Mass Function**

```{=tex}
\begin{eqnarray*}
f(x)  & = & \left\{\begin{array}{ll} p & \mbox{ if $x$ = 1}\\ 
1-p & \mbox{ if $x$ = 0 }\\ 
0 & \mbox{ otherwise}  \end{array}\right.
\end{eqnarray*}
```
## Expected Value and Variance {style="font-size: 70%;"}

**Definition:** The expected value of a discrete random variable is defined as

$$E[X] = \sum_xxP(X=x) $$ **Definition:** The variance of a discrete random variable is defined as

$$Var[X] = E[X^2] - E[X]^2$$

**Expected value for a Bernoulli Random Variable**

-   E\[$X$\] = $(0 \times P[X=0]) + (1 \times P[X=1]) = p$

**Variance for a Bernoulli Random Variable**

-   E\[$X^2$\] = $(0^2 \times p(0)) + (1^2 \times p(1)) = p$

-   Var($X$) = E\[$X^2$\] - $(E[X])^2 = p - p^2 = p(1-p)$

<!-- ## Example {style="font-size: 70%;"} -->

<!-- A company organizes a raffle at an end-of-year function. There are 300 lottery tickets in total, and 50 of them are marked as winning tickets. The event A of interest is “ticket wins” (coded as X = 1), and the probability p of having a winning ticket is a apriori (i.e. before any lottery ticket has been drawn) 1/6. -->

<!-- -   $P(X = 1) = \frac{50}{300} = \frac{1}{6} = p$ and -->

<!-- -   $P(X = 0) = \frac{250}{300} = \frac{5}{6} = 1-p$ -->

<!-- -   $E[X] = \frac{1}{6}$ -->

<!-- -   $Var[X] = \frac{1}{6} \times \frac{5}{6} = \frac{5}{36}$ -->

## The Binomial Distribution {style="font-size: 70%;"}

If $n$ Bernoulli trials all with probability $p$ are performed independently, then $X$, the number of successes out of the $n$ trials is said to be a binomial random variable with parameters $n$ and $p$.

**Example** 200 patients take a drug that has success probability 0.75. Let $X$ = \# successes. We write: $$X \sim Binomial(n=200, p=0.75).$$

**Probability mass function**

If $X \sim$ Binomial($n,p$), the possible outcomes are k = 0, 1, 2, ..., $n$.

$$P(X=k) = {n \choose k}p^k(1-p)^{n-k}$$

**Expected value**

E\[$X$\] = $np$

**Variance**

Var($X$) = $np(1-p)$

## Assumptions for a Binomial Random Variable {style="font-size: 70%;"}

For a random variable to follow a **Binomial distribution**, these key assumptions must hold:

1.  **Two Possible Outcomes per Trial**\
    Each trial has only two possible outcomes, typically labeled as **"success"** or **"failure"**.

2.  **Independence of Trials**\
    The outcome of one trial does not influence the outcome of another; trials are **independent**.

3.  **Constant Probability of Success (( p ))**\
    The probability of success remains the **same** for each trial.

4.  **Fixed Number of Trials (( n ))**\
    The process consists of a **pre-determined, fixed number** of identical trials.

## Binomial Example {style="font-size: 70%;"}

Hospital records show that of patients suffering from a certain disease, 10% die of it. Assume we monitor 50 randomly selected patients with the disease. What is the probability mass function?

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: "1-5"
#| output-location: column-fragment

n <- 50
p <- 0.9
k <- 0:n
pmf <- dbinom(k,n,p)
round(pmf,2)
```

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: "1"
#| output-location: column-fragment
plot(k,pmf, type = "h")
```

## Cumulative Distribution Function (CDF) {style="font-size: 70%;"}

::: callout-note
### Cumulative Distribution Function (CDF)

The **Cumulative Distribution Function (CDF)** gives the probability that the variable takes on a value less than or equal to a specified value.
:::

**Definition**

The CDF of a binomial random variable $X$, denoted as $F(x)$, is given by:

$$
F(x) = P(X \leq x) = \sum_{k=0}^{\lfloor x \rfloor} P(X = k) = \sum_{k=0}^{\lfloor x \rfloor} \binom{n}{k} p^k (1 - p)^{n - k}
$$

<!-- ### Example -->

<!-- For a binomial random variable \( X \) with \( n = 10 \) trials and a success probability \( p = 0.3 \): -->

<!-- - To find \( F(3) \), which is the probability of getting **3 or fewer successes**: -->

<!-- \[ -->

<!-- F(3) = P(X \leq 3) = \sum_{k=0}^{3} \binom{10}{k} (0.3)^k (0.7)^{10-k} -->

<!-- \] -->

<!-- This CDF will give the cumulative probability of observing 0, 1, 2, or 3 successes in 10 trials. -->

**Key Points**

-   The CDF is a non-decreasing function that ranges from 0 to 1.
-   As $x$ increases, $F(x)$ approaches 1.
-   The CDF can be used to calculate probabilities for a range of outcomes and is useful in hypothesis testing and statistical inference.

## Binomial Example {style="font-size: 70%;"}

Hospital records show that of patients suffering from a certain disease, 10% die of it. Assume we monitor 50 randomly selected patients with the disease. What is the cumulative distribution function?

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: ""
#| output-location: column-fragment

n <- 50
p <- 0.9
k <- 0:n
cdf <- pbinom(k,n,p)
round(cdf,2)
```

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: "1"
#| output-location: column-fragment
plot(k,cdf)
```

## Binomial Example {style="font-size: 70%;"}

What is the probability that of 50 randomly selected patients, exactly 40 will recover?

```{r, echo = TRUE, include=TRUE}
dbinom(x = 40, size = 50, prob = 0.9)
```

What is the probability that of 50 randomly selected patients, 45 or more will recover?

```{r, echo = TRUE, include=TRUE}
1 - pbinom(q = 44,size = 50, prob = 0.9)
```

What is the 80th percentile of this distribution?

```{r, echo = TRUE, include=TRUE}
qbinom(p = 0.8,size = 50, prob = 0.9)
```

## The Poisson Distribution {style="font-size: 70%;"}

The Poisson distribution is a probability model for count data. You can use a Poisson distribution to predict or explain the number of events occurring within a given interval of time or space

**Example**

Suppose the average number road deaths per month is estimated to be 12. The we say X, the number of road deaths per months has a Poisson distribution with mean $\lambda = 12$.

$$X \sim Poisson(\lambda = 12).$$

**Probability Mass Function**

$$P(X=k) = \frac{e^{-\lambda}\lambda^k}{k!}, \hspace{1em} k = 1,2, \ldots$$

**Expected value and variance**

-   $E[X] = \lambda$, $Var(X) = \lambda$

## Example {style="font-size: 70%;"}

Suppose a country experiences 4 tropical storms on average per year. What is the probability mass function?

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: ""
#| output-location: column-fragment
n <- 40
lambda <- 4
k <- 0:n
pmf <- dpois(k,lambda)
round(pmf,2)
```

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: ""
#| output-location: column-fragment

plot(k,pmf, type = "h")
```

## Cumulative Distribution Function (CDF) {style="font-size: 70%;"}

### Definition

For a Poisson random variable $X$ that represents the number of events occurring in a fixed interval of time or space, given an average rate of occurrence $\lambda$, the CDF is defined as follows:

$$
F(x) = P(X \leq x) = \sum_{k=0}^{\lfloor x \rfloor} P(X = k) = \sum_{k=0}^{\lfloor x \rfloor} \frac{e^{-\lambda} \lambda^k}{k!}
$$

### Key Points

-   The CDF is a non-decreasing function ranging from 0 to 1.
-   As $x$ increases, $F(x)$ approaches 1.
-   The CDF can be used to determine probabilities for various outcomes and is useful in statistical analysis, such as hypothesis testing and confidence intervals for count data.

## Poisson Example {style="font-size: 70%;"}

Suppose a country experiences 4 tropical storms on average per year. What is the cumulative distribution function?

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: ""
#| output-location: column-fragment

n <- 40
lambda <- 4
k <- 0:n
cdf <- ppois(k, lambda)
round(cdf,2)
```

```{r, echo = TRUE, include=TRUE}
#| code-line-numbers: ""
#| output-location: column-fragment
plot(k,cdf)
```

## Poisson Example {style="font-size: 70%;"}

What is the probability the country experiences 6 tropical storms in a given year.

```{r, echo = TRUE, include=TRUE}
dpois(x = 6,lambda)
```

What is the probability more than 4 storms will occur?

```{r, echo = TRUE, include=TRUE}
1- ppois(q = 4,lambda)
```

What is the 80th percentile of this distribution?

```{r, echo = TRUE, include=TRUE}
qpois(p = 0.8,lambda)
```
