---
title: "ST201 Tutorial Sheet 4 "
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ECHO = FALSE
INCLUDE = FALSE
```

## Instructions

-   Attempt the questions below before your tutorial in the week beginning 2nd December 2024.

**Exercise 1**

Suppose the average time between checkouts for a supermarket cashier is three minutes. Let X be a random variable describing the the time between checkouts and assume X follows an exponential distribution.

<!-- To answer probability questions you need to consider what important information you can take from above. -->

<!--   - The average time between checkouts is 3 minutes -->

<!--   - We can infer from this that the rate of checkouts per minute is 1/3 -->

<!-- So if we say that X is the time between checkouts, where -->

<!-- $$X \sim exponential(rate = 1/3)$$ -->

<!-- then we can find probabilities and percentiles using the `pexp()` and `qexp()` functions. We can simulate from this distribution using `rexp()` as shown in part (a) below. -->

(a) What is the value of the parameter $\lambda$ in the exponential distribution?

(b) Use `rexp()` to simulate 10,000 values from this distribution and plot the density curve.

```{r,  echo = ECHO, include = INCLUDE}
exp_samps <- rexp(10000,
                  rate = 1/3)
plot(density(exp_samps))
```

(b) Calculate the probability of the time between checkouts being $\leq$ 2 minutes.

```{r, echo = ECHO, include = INCLUDE}
pexp(2)
```

(c) What is the 80th percentile for this distribution?

```{r, echo = ECHO, include = INCLUDE}
qexp(0.8,
     rate = 1/3)
```

------------------------------------------------------------------------

**Exercise 2**

Suppose a Normal Distribution has $\mu$ = 0 and $\sigma = 1$.

(a) Use `rnorm()` to simulate 10,000 values from this distribution and plot the density curve.

```{r, echo = ECHO, include = INCLUDE}
norm_samps <- rnorm(10000,
                    mean = 0,
                    sd = 1)

plot(density(norm_samps))
```

(b) Using the `pnorm()` function, calculate the area below -1, above 2, between -1 and 2.

```{r, echo = ECHO, include = INCLUDE}
# below -1
pnorm(-1)

# above 2
1-pnorm(1)

# between -1 and 2
pnorm(2) - pnorm(-1)
```

(c) Plot the cumulative probability distribution function for the sequence of values between -4 and 4 at intervals of 0.1.

```{r, echo = ECHO, include = INCLUDE}

x <- seq(-4,4,by = 0.1)

cdf <- pnorm(x)

plot(x,cdf)
```

(d) Using `qnorm()`, what is the median of the distribution? The upper and lower quartiles? Find the 2.5th and 97.5th percentile values.

```{r,echo = ECHO, include = INCLUDE}
# median (note it's 0, same as the mean)
qnorm(0.5)
# lower quartile
qnorm(0.25)
# upper quartile
qnorm(0.75)
# 2.5th percentile
qnorm(0.025)
# 97.5th percentile
qnorm(0.975)

### notice the symmetry in the above answers

```

(e) How do your answers in (d) compare to answers you get when you apply the `quantile()` function to your simulated values from (a)?

```{r,echo = ECHO, include = INCLUDE}
# median
quantile(norm_samps, probs = 0.5)
# lower quantile
quantile(norm_samps, probs = 0.25)
# upper quantile
quantile(norm_samps, probs = 0.75)
# 2.5th percentile
quantile(norm_samps, probs = 0.025)
# 97.5th percentile
quantile(norm_samps, probs = 0.975)
```

------------------------------------------------------------------------

**Exercise 3**

(a) IQ tests are designed so that the average IQ is 100, the standard deviation of IQ scores is 15, and the distribution of IQ scores is normal. Run an experiment. Sample N = 5 people's IQ scores (using the population characteristics). Check the mean IQ for your sample.

(b) Now replicate the experiment. That is, randomly sample 5 new people and measure their IQ. Check the mean IQ for your sample.

(c) Repeat the experiment 1000 times and store the means in an object called `IQ_means` (note we'll see a similar example in Week9)

(d) Plot the distributions of sample means using a histogram. Overlay a density plot that reflects the population distribution. How does the distribution of the sample means compare to the population distribution for IQ scores?

(e) If you sampled N = 100 IQ scores instead of N = 5, how would this impact the sampling distributions?

------------------------------------------------------------------------

**Exercise 4**

Suppose we get 40 students from Maynooth campus to sit through an IQ test. The test results can be found in IQ_dat.rda in the ST201 folder on the R studio server. Based on this sample, what is the 95% confidence interval for the population mean IQ?

------------------------------------------------------------------------
